---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Package loading, unzip and load the .csv file.
```{r Package, load and process, message=FALSE, warning=FALSE}
library(lubridate)
library(dplyr)
library(lattice)
Sys.setlocale("LC_TIME", "English")
unzip("activity.zip")
act<-read.csv("activity.csv")
act$date<-ymd(act$date)
str(act)
```


## What is mean total number of steps taken per day?

### 1. Total number of steps taken per day:
```{r sum per day}
sumPerDay<-tapply(act$steps,act$date,sum,na.rm=T)
sumPerDay
```

### 2. Histogram of the total number of steps in each day:
```{r histogram}
hist(sumPerDay,xlab = "Total number of steps per day",main = "Histogram of total number of steps per day")
```

### 3. Mean and median of the total number of steps taken per day:

#### 3.1. Mean per day:
```{r mean per day}
meanPerDay<-tapply(act$steps,act$date,mean,na.rm=T)
meanPerDay
```

#### 3.2. Median per day:
```{r median per day}
medianPerDay<-tapply(act$steps,act$date,median,na.rm=T)
medianPerDay
```


## What is the average daily activity pattern?

### 1. Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r mean per interval}
meanPerInterval<-tapply(act$steps,act$interval,mean,na.rm=T)
plot(names(meanPerInterval),meanPerInterval,type="l",xlab = "5-minute interval",ylab="Average number of steps across all days",main = "Time series plot daily pattern",col="blue")
```

### 2. The 5-minute interval that contains the maximum number of steps on average across all the days in the dataset is:
```{r max mean per interval}
meanPerInterval<-data.frame(names(meanPerInterval),meanPerInterval)
subset(meanPerInterval$names.meanPerInterval.,meanPerInterval$meanPerInterval==max(meanPerInterval$meanPerInterval))
```


## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset
```{r NA rows quantity}
naRows<-sum(is.na(act))
naRows
```

### 2. Devise a strategy for filling in all of the missing values in the dataset
The missing values will be filled with the mean of the 5 minute interval across all days.

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r dataset with filled NA}
stepsFilled<-rep(0,nrow(act))
meanPerIntervalRep<-rep(meanPerInterval$meanPerInterval,nrow(act))
for (i in 1:nrow(act)) {
    if(is.na(act[i,1])) {
        stepsFilled[i]<-meanPerIntervalRep[i]
    }
    else {
        stepsFilled[i]<-act[i,1]
    }
}
actFilled<-data.frame(stepsFilled,act$date,act$interval)
names(actFilled)<-c("steps","date","interval")
str(actFilled)
```

### 4. Calculations with the missing data filled

#### 4.1 Histogram of the total number of steps taken each day
```{r histogram of na filled}
sumPerDayFilled<-tapply(actFilled$steps,actFilled$date,sum)
hist(sumPerDayFilled,xlab = "Total number of steps per day",main = "Histogram of total number of steps per day with missing values filled")
```

#### 4.2 Mean total number of steps taken per day
```{r mean filled}
meanPerDayFilled<-tapply(actFilled$steps,actFilled$date,mean)
meanPerDayFilled
```

#### 4.3 Median total number of steps taken per day. 
```{r median filled}
medianPerDayFilled<-tapply(actFilled$steps,actFilled$date,median)
medianPerDayFilled
```

#### 4.4 Do these values differ from the estimates from the first part of the assignment? 
Yes, the values are different.

#### 4.5 What is the impact of imputing missing data on the estimates of the total daily number of steps?
There is an important impact specially in the range from 0 to 5000.


## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r weekday weekend variable}
weekendDay<-c("Saturday","Sunday")
dayType<-function(x){
    if (weekdays(x) %in% weekendDay) return("weekend")
    else return("weekday")
}
actFilled$dayType<-as.factor(sapply(actFilled$date,dayType))
str(actFilled)
```

### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r panel plot}
actFilled$intervalDayType<-with(actFilled,paste(dayType,interval,sep="_"))
meanPerIntervalDayTypeFilled<-tapply(actFilled$steps,actFilled$intervalDayType,mean)
dayTypeInterval<-strsplit(names(meanPerIntervalDayTypeFilled),"_")
dayType<-sapply(dayTypeInterval,function(x) x[1])
interval<-sapply(dayTypeInterval,function(x) x[2])
byWd<-data.frame(as.integer(interval),meanPerIntervalDayTypeFilled,as.factor(dayType))
names(byWd)<-c("Interval","NumberOfSteps","dayType")
byWd<-arrange(byWd,Interval)
xyplot(NumberOfSteps~Interval | dayType,data=byWd,layout=c(1,2),type="l",xlab = "5-minute interval",ylab="Average number of steps across all days", main="Average number of steps by type of day")
```

We can see that on weekday we got a higher pick than on weekend but during weekend we see more activity during the day than on weekday